Errata

### 2025-04-07 tex/Manuscript-3.tex
Refactor code structure for improved readability and maintainability

Moisture and temperature changes notably impact soil microorganisms, particularly in arid and semiarid environments. Increased or decreased water availability affects gene copy numbers \citep{Barnard2013}, respiration and productivity \citep{Armstrong2016}, biomass \citep{Na2019, Ouyang2020}, diversity, and community composition \citep{Maestre2015, StovicekB2017}, due to alterations in soil properties (such as soil organic carbon content) and plant biomass. Key enzymatic activities in the carbon cycle, such as cellobiohydrolase and \textbeta-glucosidase \citep{Ouyang2020}, and in the phosphorous metabolism, such as acid phosphatase \citep{Na2019} also respond to water availability. However, the microbial response depends on the intensity and frequency of water pulses and ambient temperature, highlighting the complexity of bacterial responses \citep{StovicekA2017}. For instance, sudden and substantial water input in hyperarid regions can harm many soil microbial species and decrease diversity \citep{AzuaBustos2018}, while in other studies in the desert, it did not affect diversity \citep{Armstrong2016}. As our understanding of the direction and magnitude of microbial feedback to climate change continues to advance, it becomes increasingly important to investigate their implications for soil formation.

@@

We collected soil samples from two protected areas, Pan de Az{\'u}car National Park (PA) and Santa Gracia Natural Reserve (SG), located in the Chilean Coastal Cordillera (\cref{fig:figure_label}, Supplementary \cref{fig:supp_figure_label}). 
The lithology at both sites is granitoid, ensuring similar parent material for soil formation and a meaningful comparison of the obtained results \citep{Bernhard2018, Oeser2018}. 
The mean annual precipitation and mean annual temperature are \SI{12}{\milli\metre} and \SI{16.8}{\degreeCelsius} in PA and \SI{66}{\milli\metre} and \SI{13.7}{\degreeCelsius} in SG, respectively \citep{Munoz2007}. 
These climatic differences result in distinct soil moisture and temperature regimes. 
PA is arid with almost no precipitation and mostly endorheic water resources. 
In contrast, SG is in a semiarid zone with coastal fog influence, annual rainfall variation, and winter rain \citep{Bernhard2018}.

@@

The field sampling campaign took place in March 2019, at the end of the austral summer season. The coordinates and elevations for PA were \ang{26;06;35}S, \ang{70;32;57}W, and \(\sim\)\SI{336}{\meter} elevation, while for SG, they were \ang{29;45;15}S, \ang{71;10;3}W, and \(\sim\)\SI{720}{\meter} elevation. Soil samples were taken from five plots devoid of vegetation and two soil horizons (A and B) for each site (Supplementary \cref{tab:supp_table_label}). In PA, the A horizon (PA - Ahor) ranged from \SIrange{0}{2}{\centi\metre}, and the B horizon (PA - Bhor) ranged from \SIrange{2}{25}{\centi\metre}. Similarly, in SG, the A horizon (SG - Ahor) ranged from \SIrange{0}{3}{\centi\metre}, and the B horizon (SG - Bhor) ranged from \SIrange{3}{40}{\centi\metre}. The soil structure characteristics varied between horizons. PA - Ahor exhibited a cemented crust, while PA - Bhor had a few round aggregates between coarse material, single grain dominating in the fine fraction, and gypsum concentrated in one of the plots. SG - Ahor was characterized by weak aggregates, and SG - Bhor ranged from weak to strong and stable aggregates. Fresh samples from each plot and horizon separately were homogenized through a \SI{2}{\milli\metre} sieve and stored at \SI{4}{\degreeCelsius} until use.

@@

For the manipulation experiment setup, soil material from five plots representing top and subsoil was pooled and homogenized. The experiment included 144 microcosms representing two sites (PA and SG), two soil horizons (Ahor and Bhor), four treatments, three sampling times, and three biological replicates for each combination (Supplementary \cref{fig:supp_figure_s2_label}). In addition, three original, untreated samples were collected as baseline measurements for each site and horizon (T0), totaling 156 samples. The treatments consisted of a successional gradient of soil development as follows: (i) in situ samples with indigenous microorganisms but without BSCs or plants, (ii) in situ samples with naturally growing BSCs to assess the impact of microorganisms at a later stage of soil development, (iii) in situ samples with \textit{Helenium aromaticum} (Hook.) L.H. Bailey (pioneer plant, described below) to simulate later stages of soil development and (iv) abiotic controls based on soil sterilized samples with \ce{^{60}Co} as a source of gamma rays and a dose of \SI{50}{\kilo Gy} (BGS Beta–Gamma–Service GmbH and Co. KG, Wiehl, Germany) to exclude the impact of indigenous microorganisms. \textit{H. aromaticum} is a native vascular annual herb widely distributed in Chile, known as a pioneer plant that colonizes disturbed soils, forming seasonal grasslands \citep{GomezGonzalez2011}. For plant preparation, the seeds were sterilized using 20\% v/v sodium hypochlorite and 0.05\% Tween 20, washed three times with autoclaved MiliQ water, and placed in sterile petri dishes with absorbent paper for germination. The germinated seeds were subsequently transferred to the microcosms. BSCs occurred naturally in the soil, whereas in situ samples with microorganisms were covered with a black breathable cloth to inhibit their growth.

@@

The microcosms represent a successional gradient of soil development divided into three biological treatments (\cref{fig:figure_1_label}, detailed in Supplementary \cref{fig:supp_figure_s3_label}): soil with \textit{in situ} microorganisms (\textit{in situ}), soil with \textit{in situ} microorganisms and BSCs (BSC), and soil with \textit{in situ} microorganisms and plants (plant). The \textit{in situ} treatment had no plant growth or BSC development, similar to the sterile treatment. In the BSC treatment, varying types and coverage were observed among sites and horizons. Specifically, a leaden BSC developed in PA, while a greenish BSC formed in SG in both horizons. The A horizons of both sites showed nearly complete coverage of the microcosms with BSCs. In contrast, the B horizons showed patchy coverage without forming a compact or extended layer of BSC on the surface. Regarding the plant treatments, similar patterns were observed. In PA, the roots were light brown, and the biomass (both below- and aboveground) and vegetative development were comparatively lower than in SG. In contrast, SG showed dark brown roots, higher biomass, and reached the flowering stage in both horizons. In the PA-Ahor, plants with numerous lateral roots ranged from \SIrange{2}{3}{\centi\metre} in height. Conversely, in PA–Bhor, plants were less than \SI{5}{\milli\metre} tall, had a short taproot with few lateral roots, and experienced a mortality rate of over \SI{50}{\percent} after seven weeks. In SG–Ahor, flowering plants with heights ranging from \SIrange{11}{14}{\centi\metre} were observed. These plants had a \SI{7}{\centi\metre} long taproot and numerous lateral roots covering completely the microcosms. On the other hand, SG–Bhor exhibited plants ranging from \SIrange{1}{3}{\centi\metre} in height, with flower buds and roots reaching \SI{5}{\centi\metre} in length with fewer lateral roots than SG–Ahor. After removing the plants at 12 weeks, no visual differences were observed in the roots between weeks 12 and 16, which remained compact and defined with no signs of degradation. Given the intact nature of the roots during this period, it is considered that these conditions did not impact the short-term degradation of this organic material. Overall, the differences between treatments were less pronounced in the B horizons than in the A horizons, indicating more favorable conditions for BSC and plant development in the A horizons.

@@

Soil samples were analyzed to determine the mass distribution of macro- and microaggregates and the concentrations of C\textsubscript{t} and N\textsubscript{t} in each soil aggregate size (Supplementary \cref{tab:supp_table_s2_label}). No significant differences in aggregate size distribution were found between the study sites (PA--SG; \(n = 54\)), with average proportions of \SI{67}{\percent} macroaggregates, \SI{27}{\percent} large microaggregates, and \SI{7}{\percent} small microaggregates. However, significant differences existed between horizons within each site (Ahor--Bhor; \(n = 27\), \(p < 0.01\)). In both sites, the proportion of macroaggregates was highest in the B horizon, while microaggregates dominated the A horizon. Specifically, macroaggregates mean values were \SI{54}{\percent} in PA--Ahor, \SI{81}{\percent} in PA--Bhor, \SI{63}{\percent} in SG--Ahor, and \SI{70}{\percent} in SG--Bhor. For microaggregates, the means were \SI{23}{\percent} in PA--Ahor, \SI{9}{\percent} in PA--Bhor, \SI{19}{\percent} in SG--Ahor, and \SI{15}{\percent} in SG--Bhor (The full dataset is available in Supplementary \cref{tab:supp_table_s2_label}). Significant differences in the concentration of C\textsubscript{t} and N\textsubscript{t} were observed between sites and horizons (\(p < 0.01\)). C\textsubscript{t} was higher in macro- and large microaggregates of PA (mean \SI{1.2}{\percent}) than in SG (mean \SI{0.7}{\percent}), while N\textsubscript{t} was significantly higher in macro- and microaggregates of SG (mean \SI{0.08}{\percent}) than in PA (mean \SI{0.03}{\percent}). Analysis between horizons showed that macro- and microaggregates of PA--Bhor had a higher C\textsubscript{t} mean of \SI{2}{\percent} than PA--Ahor’s \SI{0.8}{\percent}. Conversely, SG--Ahor’s macro- and large microaggregates had a higher C\textsubscript{t} mean of \SI{1.1}{\percent} than SG--Bhor’s \SI{0.8}{\percent}. In terms of N\textsubscript{t}, higher percentages were observed in SG--Ahor (mean \SI{0.1}{\percent}) and PA--Ahor (mean \SI{0.033}{\percent}) than in SG--Bhor (mean \SI{0.07}{\percent}) and PA--Bhor (mean \SI{0.026}{\percent}).

Over time, we observed changes in aggregate size distribution as well as in C\textsubscript{t} and N\textsubscript{t} concentrations (Supplementary \cref{tab:supp_table_s2_label}). In PA--Ahor, the C\textsubscript{t} of large microaggregates decreased from \SI{0.93}{\percent} in T0 to \SI{0.83}{\percent} in T3. In PA--Bhor, the proportion of macroaggregates increased from \SI{67}{\percent} to \SI{81}{\percent}, while for large microaggregates decreased from \SI{28}{\percent} to \SI{14}{\percent} between T0 and T3. Consequently, MWD increased from \SI{0.8}{\milli\metre} in T0 to \SI{0.94}{\milli\metre} in T3. In SG--Ahor, no marked contrast was observed from T0 to T3, while in SG--Bhor, the percentage of C\textsubscript{t} in macroaggregates increased from \SI{0.2}{\percent} to \SI{0.3}{\percent} over time. Regarding differences between treatments, microorganism-plant treatments showed a noticeable increase in macroaggregates and, thus, in MWD compared to the sterile treatments at T3 (\cref{fig:figure_2_label}). The MWD values for sterile compared to plant treatments were as follows: PA--Ahor (mean \SI{0.7}{mm} vs. \SI{0.78}{\milli\metre}), PA--Bhor (mean \SI{0.91}{mm} vs. \SI{0.95}{\milli\metre}), SG--Ahor (mean \SI{0.75}{mm} vs. \SI{0.87}{\milli\metre}), and SG--Bhor (mean \SI{0.87}{mm} vs. \SI{0.89}{\milli\metre}).

Regarding TOC and N\textsubscript{bulk} (without aggregate size separation) in the bulk soil samples, SG exhibited significantly higher levels (\SI{0.7}{\percent} and \SI{0.06}{\percent}, respectively) than PA (\SI{0.1}{\percent} and \SI{0.02}{\percent}, respectively). When considering the horizons, both N\textsubscript{bulk} and TOC were significantly higher in PA--Ahor (\SI{0.03}{\percent} and \SI{0.14}{\percent}, respectively) and SG--Ahor (\SI{0.08}{\percent} and \SI{0.9}{\percent}, respectively) than PA--Bhor (\SI{0.01}{\percent} and \SI{0.06}{\percent}, respectively) and SG--Bhor (\SI{0.04}{\percent} and \SI{0.46}{\percent}, respectively). Over time, a notable increase in TOC was observed only in SG--Bhor, rising from \SI{0.32}{\percent} in T0 to \SI{0.48}{\percent} in T3. No clear trends were identified for the treatments (\cref{fig:figure_2_label}, Supplementary \cref{tab:supp_table_s2_label}).

@@

The results revealed significant differences in soil pH, EC, and ion concentrations among the studied sites and horizons (\(p < 0.01\); \cref{fig:figure_2_label}, Supplementary \cref{tab:supp_table_s2_label}). Specifically, in PA, higher values were observed for pH (mean 7.8), EC (mean \SI{1272}{\micro\siemens\per\centi\metre}), sodium (mean \SI{93.4}{\milli\gram\per\liter}), chloride (mean \SI{160.2}{\milli\gram\per\liter}), and sulfate (mean \SI{783.6}{\milli\gram\per\liter}) compared to SG, where the respective mean values were 6.5, \SI{51.8}{\micro\siemens\per\centi\metre}, \SI{5.5}{\milli\gram\per\liter}, \SI{6.2}{\milli\gram\per\liter}, and \SI{1.4}{\milli\gram\per\liter}. Furthermore, significant differences in soil parameters were found between soil horizons. PA--Bhor exhibited higher EC (mean \SI{2186}{\micro\siemens\per\centi\metre}), sodium (mean \SI{141}{\milli\gram\per\liter}), chloride (mean \SI{246.7}{\milli\gram\per\liter}), calcium (mean \SI{630}{\milli\gram\per\liter}), and sulfate (mean \SI{1532.6}{\milli\gram\per\liter}) concentrations than PA--Ahor, where the respective mean values were \SI{359}{\micro\siemens\per\centi\metre}, \SI{45.8}{\milli\gram\per\liter}, \SI{73.8}{\milli\gram\per\liter}, \SI{12.5}{\milli\gram\per\liter}, and \SI{14.8}{\milli\gram\per\liter}. In SG, higher pH and EC were observed in the A horizon (mean 6.6 and \SI{63}{\micro\siemens\per\centi\metre}, respectively) than in the B horizon (mean 6.4 and \SI{41}{\micro\siemens\per\centi\metre}, respectively). Additionally, changes over time were observed, with substantial decreases of EC (from \num{1809} to \num{103.7}\,\si{\micro\siemens\per\centi\metre}), sodium (from \num{108.8} to \SI{32.7}{\milli\gram\per\liter}), and chloride (from \num{265.3} to \SI{48.2}{\milli\gram\per\liter}) concentrations from T0 to T3 in PA--Ahor. PA--Bhor also showed a marked decrease in sodium and chloride concentration from \num{250.2} and \SI{522.6}{\milli\gram\per\liter} to \num{109} and \SI{164.5}{\milli\gram\per\liter}, respectively. 
No remarkable changes were found for the horizons of SG over time, as well as for treatments in PA or SG.

@@

The bacterial gene copy numbers are shown in \cref{fig:figure_3_label} and Supplementary \cref{tab:supp_table_s3_label}. A significantly higher gene copy number for bacteria was observed in SG, ranging from \num{e7} to \num{e8}~gene copies\,\si{\gram^{-1}} soil, compared to PA, ranging from \num{e5} to \num{e7}~gene copies\,\si{\gram^{-1}} soil. Comparison between soil horizons showed significantly higher gene copy numbers in the A horizon for both sites. The mean of gene copies\,\si{\gram^{-1}} soil for PA--Ahor was \num{4.5e7} and for PA--Bhor \num{1.2e6}, while for SG--Ahor was \num{3.9e8} and for SG--Bhor \num{9.2e7}. Considering the times evaluated, both horizons in PA showed an increase in the gene copy numbers from T0 (mean of gene copies\,\si{\gram^{-1}} soil for Ahor \num{1.9e7} and Bhor \num{8.3e5}) to T3 (mean of gene copies\,\si{\gram^{-1}} soil for Ahor \num{4.7e7} and Bhor \num{1.1e6}). In contrast, SG decreased the gene copy number over time in both horizons. The mean of gene copy\,\si{\gram^{-1}} soil for SG--Ahor decreased from \num{5e8} in T0 to \num{2.9e8} in T3, while SG--Bhor reduced from \num{1.8e8} in T0 to \num{8.2e7} in T3. There were no discernible patterns between treatments. The gene copy numbers correlated positively with N\textsubscript{t} in all aggregate classes, N\textsubscript{bulk} and TOC (\(R > 0.7\)), and negatively with pH (\(R = 0.6\)).

@@

Sequencing analysis of 116 soil samples yielded 4.3 million raw reads, of which 3.1 million (\SI{72.7}{\percent} of the total) were retained after filtering, merging, and removing chimera (Supplementary \cref{tab:supp_table_s4_label}). These reads were assigned to bacteria (\SI{98.5}{\percent}), archaea (\SI{1.3}{\percent}), or unclassified (\SI{0.2}{\percent}) groups. Since the percentage of archaea was low (with a mean of 594 reads per sample), only bacterial data were analyzed. The number of reads per bacterial sample ranged from \num{12178} to \num{275292}, with a mean of \num{90155} reads per sample. Considering the minimum read number, the samples were rarified to 12,000 reads. The final dataset contained \num{24465} amplicon sequence variants (ASVs) and \num{1108} putative genera.

This study evaluated the alpha-diversity indices, including the Shannon index, observed species, and evenness (\cref{fig:figure_3_label}, Supplementary \cref{tab:supp_table_s3_label}). The results revealed significant differences between PA and SG, with SG exhibiting higher values for all three indices. The Shannon index ranged from 4.1 to 6.3 in PA, while SG showed values ranging from 6.7 to 7.7. A similar trend was noticed for the observed species, with PA ranging from 550 to 1,754 and SG ranging from 2,075 to 4,029. For evenness, values varied between 0.87 to 0.93 in SG and 0.64 to 0.88 in PA. For the soil horizons, only the observed species values showed significant differences. PA--Ahor had a higher mean of 1,215 compared to 1,050 in PA--Bhor, whereas SG--Bhor showed a mean of 3,541 compared to 3,389 in SG--Ahor. Over time, greater changes were observed in both horizons of PA. Samples analyzed at T0 in PA--Ahor showed a marked decrease in the Shannon index and evenness at T2. The mean Shannon index decreased from 5.9 to 5.4, while evenness decreased from a mean of 0.83 to 0.76 in the same period. In PA--Bhor, the Shannon index, evenness, and observed species prominently reduced from T0 to T3. The mean Shannon index decreased from 6.2 to 5.1, evenness from 0.84 to 0.76, and the observed species mean varied from 1,568 in T0 to 892 in T3. No robust differences were observed between time or treatments in SG. However, in PA--Bhor, the Shannon index showed considerable disparities between treatments at T3, with the lowest values observed in the \textit{in situ} treatment (4.7), followed by plants (5.2) and BSC (5.6).

>>>>

Moisture and temperature changes notably impact soil microorganisms, particularly in arid and semiarid environments. Increased or decreased water availability affects gene copy numbers \citep{Barnard2013}, respiration and productivity \citep{Armstrong2016}, biomass \citep{Na2019, Ouyang2020}, diversity, and community composition \citep{Maestre2015, StovicekB2017}, due to alterations in soil properties (such as soil organic carbon content) and plant biomass. Key enzymatic activities in the carbon cycle, such as cellobiohydrolase and $\beta$-glucosidase \citep{Ouyang2020}, and in the phosphorous metabolism, such as acid phosphatase \citep{Na2019} also respond to water availability. However, the microbial response depends on the intensity and frequency of water pulses and ambient temperature, highlighting the complexity of bacterial responses \citep{StovicekA2017}. For instance, sudden and substantial water input in hyperarid regions can harm many soil microbial species and decrease diversity \citep{AzuaBustos2018}, while in other studies in the desert, it did not affect diversity \citep{Armstrong2016}. As our understanding of the direction and magnitude of microbial feedback to climate change continues to advance, it becomes increasingly important to investigate their implications for soil formation.

@@

We collected soil samples from two protected areas, Pan de Az{\'u}car National Park (PA) and Santa Gracia Natural Reserve (SG), located in the Chilean Coastal Cordillera (\ref{fig:M3-F1}). 
The lithology at both sites is granitoid, ensuring similar parent material for soil formation and a meaningful comparison of the obtained results \citep{Bernhard2018, Oeser2018}. 
The mean annual precipitation and mean annual temperature are \SI{12}{\milli\metre} and \SI{16.8}{\degreeCelsius} in PA and \SI{66}{\milli\metre} and \SI{13.7}{\degreeCelsius} in SG, respectively \citep{Munoz2007}. 
These climatic differences result in distinct soil moisture and temperature regimes. 
PA is arid with almost no precipitation and mostly endorheic water resources. 
In contrast, SG is in a semiarid zone with coastal fog influence, annual rainfall variation, and winter rain \citep{Bernhard2018}.

@@

The field sampling campaign took place in March 2019, at the end of the austral summer season. The coordinates and elevations for PA were \ang{26;06;35}S, \ang{70;32;57}W, and \(\sim\)\SI{336}{\meter} elevation, while for SG, they were \ang{29;45;15}S, \ang{71;10;3}W, and \(\sim\)\SI{720}{\meter} elevation. Soil samples were taken from five plots devoid of vegetation and two soil horizons (A and B) for each site. In PA, the A horizon (PA - Ahor) ranged from \SIrange{0}{2}{\centi\metre}, and the B horizon (PA - Bhor) ranged from \SIrange{2}{25}{\centi\metre}. Similarly, in SG, the A horizon (SG - Ahor) ranged from \SIrange{0}{3}{\centi\metre}, and the B horizon (SG - Bhor) ranged from \SIrange{3}{40}{\centi\metre}. The soil structure characteristics varied between horizons. PA - Ahor exhibited a cemented crust, while PA - Bhor had a few round aggregates between coarse material, single grain dominating in the fine fraction, and gypsum concentrated in one of the plots. SG - Ahor was characterized by weak aggregates, and SG - Bhor ranged from weak to strong and stable aggregates. Fresh samples from each plot and horizon separately were homogenized through a \SI{2}{\milli\metre} sieve and stored at \SI{4}{\degreeCelsius} until use.

@@

For the manipulation experiment setup, soil material from five plots representing top and subsoil was pooled and homogenized. The experiment included 144 microcosms representing two sites (PA and SG), two soil horizons (Ahor and Bhor), four treatments, three sampling times, and three biological replicates for each combination. In addition, three original, untreated samples were collected as baseline measurements for each site and horizon (T0), totaling 156 samples. The treatments consisted of a successional gradient of soil development as follows: (i) in situ samples with indigenous microorganisms but without BSCs or plants, (ii) in situ samples with naturally growing BSCs to assess the impact of microorganisms at a later stage of soil development, (iii) in situ samples with \textit{Helenium aromaticum} (Hook.) L.H. Bailey (pioneer plant, described below) to simulate later stages of soil development and (iv) abiotic controls based on soil sterilized samples with \ce{^{60}Co} as a source of gamma rays and a dose of \SI{50}{\kilo Gy} (BGS Beta–Gamma–Service GmbH and Co. KG, Wiehl, Germany) to exclude the impact of indigenous microorganisms. \textit{H. aromaticum} is a native vascular annual herb widely distributed in Chile, known as a pioneer plant that colonizes disturbed soils, forming seasonal grasslands \citep{GomezGonzalez2011}. For plant preparation, the seeds were sterilized using 20\% v/v sodium hypochlorite and 0.05\% Tween 20, washed three times with autoclaved MiliQ water, and placed in sterile petri dishes with absorbent paper for germination. The germinated seeds were subsequently transferred to the microcosms. BSCs occurred naturally in the soil, whereas in situ samples with microorganisms were covered with a black breathable cloth to inhibit their growth.

@@

The microcosms represent a successional gradient of soil development divided into three biological treatments (\ref{fig:M3-F1}: soil with \textit{in situ} microorganisms (\textit{in situ}), soil with \textit{in situ} microorganisms and BSCs (BSC), and soil with \textit{in situ} microorganisms and plants (plant). The \textit{in situ} treatment had no plant growth or BSC development, similar to the sterile treatment. In the BSC treatment, varying types and coverage were observed among sites and horizons. Specifically, a leaden BSC developed in PA, while a greenish BSC formed in SG in both horizons. The A horizons of both sites showed nearly complete coverage of the microcosms with BSCs. In contrast, the B horizons showed patchy coverage without forming a compact or extended layer of BSC on the surface. Regarding the plant treatments, similar patterns were observed. In PA, the roots were light brown, and the biomass (both below- and aboveground) and vegetative development were comparatively lower than in SG. In contrast, SG showed dark brown roots, higher biomass, and reached the flowering stage in both horizons. In the PA-Ahor, plants with numerous lateral roots ranged from \SIrange{2}{3}{\centi\metre} in height. Conversely, in PA–Bhor, plants were less than \SI{5}{\milli\metre} tall, had a short taproot with few lateral roots, and experienced a mortality rate of over \SI{50}{\percent} after seven weeks. In SG–Ahor, flowering plants with heights ranging from \SIrange{11}{14}{\centi\metre} were observed. These plants had a \SI{7}{\centi\metre} long taproot and numerous lateral roots covering completely the microcosms. On the other hand, SG–Bhor exhibited plants ranging from \SIrange{1}{3}{\centi\metre} in height, with flower buds and roots reaching \SI{5}{\centi\metre} in length with fewer lateral roots than SG–Ahor. After removing the plants at 12 weeks, no visual differences were observed in the roots between weeks 12 and 16, which remained compact and defined with no signs of degradation. Given the intact nature of the roots during this period, it is considered that these conditions did not impact the short-term degradation of this organic material. Overall, the differences between treatments were less pronounced in the B horizons than in the A horizons, indicating more favorable conditions for BSC and plant development in the A horizons.

@@

Soil samples were analyzed to determine the mass distribution of macro- and microaggregates and the concentrations of C\textsubscript{t} and N\textsubscript{t} in each soil aggregate size. No significant differences in aggregate size distribution were found between the study sites (PA--SG; \(n = 54\)), with average proportions of \SI{67}{\percent} macroaggregates, \SI{27}{\percent} large microaggregates, and \SI{7}{\percent} small microaggregates. However, significant differences existed between horizons within each site (Ahor--Bhor; \(n = 27\), \(p < 0.01\)). In both sites, the proportion of macroaggregates was highest in the B horizon, while microaggregates dominated the A horizon. Specifically, macroaggregates mean values were \SI{54}{\percent} in PA--Ahor, \SI{81}{\percent} in PA--Bhor, \SI{63}{\percent} in SG--Ahor, and \SI{70}{\percent} in SG--Bhor. For microaggregates, the means were \SI{23}{\percent} in PA--Ahor, \SI{9}{\percent} in PA--Bhor, \SI{19}{\percent} in SG--Ahor, and \SI{15}{\percent} in SG--Bhor. Significant differences in the concentration of C\textsubscript{t} and N\textsubscript{t} were observed between sites and horizons (\(p < 0.01\)). C\textsubscript{t} was higher in macro- and large microaggregates of PA (mean \SI{1.2}{\percent}) than in SG (mean \SI{0.7}{\percent}), while N\textsubscript{t} was significantly higher in macro- and microaggregates of SG (mean \SI{0.08}{\percent}) than in PA (mean \SI{0.03}{\percent}). Analysis between horizons showed that macro- and microaggregates of PA--Bhor had a higher C\textsubscript{t} mean of \SI{2}{\percent} than PA--Ahor’s \SI{0.8}{\percent}. Conversely, SG--Ahor’s macro- and large microaggregates had a higher C\textsubscript{t} mean of \SI{1.1}{\percent} than SG--Bhor’s \SI{0.8}{\percent}. In terms of N\textsubscript{t}, higher percentages were observed in SG--Ahor (mean \SI{0.1}{\percent}) and PA--Ahor (mean \SI{0.033}{\percent}) than in SG--Bhor (mean \SI{0.07}{\percent}) and PA--Bhor (mean \SI{0.026}{\percent}).

Over time, we observed changes in aggregate size distribution as well as in C\textsubscript{t} and N\textsubscript{t} concentrations. In PA--Ahor, the C\textsubscript{t} of large microaggregates decreased from \SI{0.93}{\percent} in T0 to \SI{0.83}{\percent} in T3. In PA--Bhor, the proportion of macroaggregates increased from \SI{67}{\percent} to \SI{81}{\percent}, while for large microaggregates decreased from \SI{28}{\percent} to \SI{14}{\percent} between T0 and T3. Consequently, MWD increased from \SI{0.8}{\milli\metre} in T0 to \SI{0.94}{\milli\metre} in T3. In SG--Ahor, no marked contrast was observed from T0 to T3, while in SG--Bhor, the percentage of C\textsubscript{t} in macroaggregates increased from \SI{0.2}{\percent} to \SI{0.3}{\percent} over time. Regarding differences between treatments, microorganism-plant treatments showed a noticeable increase in macroaggregates and, thus, in MWD compared to the sterile treatments at T3 (\ref{fig:M3-F2}). The MWD values for sterile compared to plant treatments were as follows: PA--Ahor (mean \SI{0.7}{mm} vs. \SI{0.78}{\milli\metre}), PA--Bhor (mean \SI{0.91}{mm} vs. \SI{0.95}{\milli\metre}), SG--Ahor (mean \SI{0.75}{mm} vs. \SI{0.87}{\milli\metre}), and SG--Bhor (mean \SI{0.87}{mm} vs. \SI{0.89}{\milli\metre}).

Regarding TOC and N\textsubscript{bulk} (without aggregate size separation) in the bulk soil samples, SG exhibited significantly higher levels (\SI{0.7}{\percent} and \SI{0.06}{\percent}, respectively) than PA (\SI{0.1}{\percent} and \SI{0.02}{\percent}, respectively). When considering the horizons, both N\textsubscript{bulk} and TOC were significantly higher in PA--Ahor (\SI{0.03}{\percent} and \SI{0.14}{\percent}, respectively) and SG--Ahor (\SI{0.08}{\percent} and \SI{0.9}{\percent}, respectively) than PA--Bhor (\SI{0.01}{\percent} and \SI{0.06}{\percent}, respectively) and SG--Bhor (\SI{0.04}{\percent} and \SI{0.46}{\percent}, respectively). Over time, a notable increase in TOC was observed only in SG--Bhor, rising from \SI{0.32}{\percent} in T0 to \SI{0.48}{\percent} in T3. No clear trends were identified for the treatments (\ref{fig:M3-F2}).

@@

The results revealed significant differences in soil pH, EC, and ion concentrations among the studied sites and horizons (\(p < 0.01\); \ref{fig:M3-F2}). Specifically, in PA, higher values were observed for pH (mean 7.8), EC (mean \SI{1272}{\micro\siemens\per\centi\metre}), sodium (mean \SI{93.4}{\milli\gram\per\liter}), chloride (mean \SI{160.2}{\milli\gram\per\liter}), and sulfate (mean \SI{783.6}{\milli\gram\per\liter}) compared to SG, where the respective mean values were 6.5, \SI{51.8}{\micro\siemens\per\centi\metre}, \SI{5.5}{\milli\gram\per\liter}, \SI{6.2}{\milli\gram\per\liter}, and \SI{1.4}{\milli\gram\per\liter}. Furthermore, significant differences in soil parameters were found between soil horizons. PA--Bhor exhibited higher EC (mean \SI{2186}{\micro\siemens\per\centi\metre}), sodium (mean \SI{141}{\milli\gram\per\liter}), chloride (mean \SI{246.7}{\milli\gram\per\liter}), calcium (mean \SI{630}{\milli\gram\per\liter}), and sulfate (mean \SI{1532.6}{\milli\gram\per\liter}) concentrations than PA--Ahor, where the respective mean values were \SI{359}{\micro\siemens\per\centi\metre}, \SI{45.8}{\milli\gram\per\liter}, \SI{73.8}{\milli\gram\per\liter}, \SI{12.5}{\milli\gram\per\liter}, and \SI{14.8}{\milli\gram\per\liter}. In SG, higher pH and EC were observed in the A horizon (mean 6.6 and \SI{63}{\micro\siemens\per\centi\metre}, respectively) than in the B horizon (mean 6.4 and \SI{41}{\micro\siemens\per\centi\metre}, respectively). Additionally, changes over time were observed, with substantial decreases of EC (from \num{1809} to \num{103.7}\,\si{\micro\siemens\per\centi\metre}), sodium (from \num{108.8} to \SI{32.7}{\milli\gram\per\liter}), and chloride (from \num{265.3} to \SI{48.2}{\milli\gram\per\liter}) concentrations from T0 to T3 in PA--Ahor. PA--Bhor also showed a marked decrease in sodium and chloride concentration from \num{250.2} and \SI{522.6}{\milli\gram\per\liter} to \num{109} and \SI{164.5}{\milli\gram\per\liter}, respectively. 
No remarkable changes were found for the horizons of SG over time, as well as for treatments in PA or SG.

@@

The bacterial gene copy numbers are shown in \ref{fig:M3-F3}. A significantly higher gene copy number for bacteria was observed in SG, ranging from \num{e7} to \num{e8}~gene copies\,\si{\gram^{-1}} soil, compared to PA, ranging from \num{e5} to \num{e7}~gene copies\,\si{\gram^{-1}} soil. Comparison between soil horizons showed significantly higher gene copy numbers in the A horizon for both sites. The mean of gene copies\,\si{\gram^{-1}} soil for PA--Ahor was \num{4.5e7} and for PA--Bhor \num{1.2e6}, while for SG--Ahor was \num{3.9e8} and for SG--Bhor \num{9.2e7}. Considering the times evaluated, both horizons in PA showed an increase in the gene copy numbers from T0 (mean of gene copies\,\si{\gram^{-1}} soil for Ahor \num{1.9e7} and Bhor \num{8.3e5}) to T3 (mean of gene copies\,\si{\gram^{-1}} soil for Ahor \num{4.7e7} and Bhor \num{1.1e6}). In contrast, SG decreased the gene copy number over time in both horizons. The mean of gene copy\,\si{\gram^{-1}} soil for SG--Ahor decreased from \num{5e8} in T0 to \num{2.9e8} in T3, while SG--Bhor reduced from \num{1.8e8} in T0 to \num{8.2e7} in T3. There were no discernible patterns between treatments. The gene copy numbers correlated positively with N\textsubscript{t} in all aggregate classes, N\textsubscript{bulk} and TOC (\(R > 0.7\)), and negatively with pH (\(R = 0.6\)).

@@

Sequencing analysis of 116 soil samples yielded 4.3 million raw reads, of which 3.1 million (\SI{72.7}{\percent} of the total) were retained after filtering, merging, and removing chimera. These reads were assigned to bacteria (\SI{98.5}{\percent}), archaea (\SI{1.3}{\percent}), or unclassified (\SI{0.2}{\percent}) groups. Since the percentage of archaea was low (with a mean of 594 reads per sample), only bacterial data were analyzed. The number of reads per bacterial sample ranged from \num{12178} to \num{275292}, with a mean of \num{90155} reads per sample. Considering the minimum read number, the samples were rarified to 12,000 reads. The final dataset contained \num{24465} amplicon sequence variants (ASVs) and \num{1108} putative genera.

This study evaluated the alpha-diversity indices, including the Shannon index, observed species, and evenness (\ref{fig:M3-F3}. The results revealed significant differences between PA and SG, with SG exhibiting higher values for all three indices. The Shannon index ranged from 4.1 to 6.3 in PA, while SG showed values ranging from 6.7 to 7.7. A similar trend was noticed for the observed species, with PA ranging from 550 to 1,754 and SG ranging from 2,075 to 4,029. For evenness, values varied between 0.87 to 0.93 in SG and 0.64 to 0.88 in PA. For the soil horizons, only the observed species values showed significant differences. PA--Ahor had a higher mean of 1,215 compared to 1,050 in PA--Bhor, whereas SG--Bhor showed a mean of 3,541 compared to 3,389 in SG--Ahor. Over time, greater changes were observed in both horizons of PA. Samples analyzed at T0 in PA--Ahor showed a marked decrease in the Shannon index and evenness at T2. The mean Shannon index decreased from 5.9 to 5.4, while evenness decreased from a mean of 0.83 to 0.76 in the same period. In PA--Bhor, the Shannon index, evenness, and observed species prominently reduced from T0 to T3. The mean Shannon index decreased from 6.2 to 5.1, evenness from 0.84 to 0.76, and the observed species mean varied from 1,568 in T0 to 892 in T3. No robust differences were observed between time or treatments in SG. However, in PA--Bhor, the Shannon index showed considerable disparities between treatments at T3, with the lowest values observed in the \textit{in situ} treatment (4.7), followed by plants (5.2) and BSC (5.6).
